---
title: "Radical Sensitivity: Visualized"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
runtime: shiny
---

### An Introduction to Fiona Apple

```{r}
library(shiny)
library(shinyjs)
library(spotifyr)
library(tidyverse)
library(ggplot2)
library(plotly)

# Fetch the data from the Spotify API
corpus <- get_playlist_audio_features("", "7LDyKnV7gitLREBYl6kVLz")

# Specify the order of albums
album_order <- c("Tidal", "When The Pawn...", "Extraordinary Machine", 
                 "The Idler Wheel Is Wiser Than the Driver of the Screw and Whipping Cords Will Serve You More Than Ropes Will Ever Do (Expanded Edition)", 
                 "Fetch The Bolt Cutters")

# Define color palette
album_colors <- c("Tidal" = "blue", 
                  "When The Pawn..." = "red", 
                  "Extraordinary Machine" = "green", 
                  "The Idler Wheel Is Wiser Than the Driver of the Screw and Whipping Cords Will Serve You More Than Ropes Will Ever Do (Expanded Edition)" = "brown", 
                  "Fetch The Bolt Cutters" = "purple")

# Function to create a plot for each album
plot_album <- function(album_name) {
  # Filter data for the specified album
  album_data <- corpus %>% filter(track.album.name == album_name)
  
  # Create the plot
  p <- ggplot(album_data, aes(x = acousticness, y = valence, size = energy, 
                              text = paste("Song:", track.name, "<br>Album:", track.album.name, "<br>Energy:", energy))) +  # Define tooltip text
    geom_point(aes(color = album_name)) +  # Set color to album
    scale_color_manual(values = album_colors, guide = FALSE) +  # Remove legend for album name
    scale_size_continuous(range = c(3, 10), guide = FALSE) +  # Set point size without legend
    theme_minimal() +
    labs(title = ifelse(album_name == "The Idler Wheel Is Wiser Than the Driver of the Screw and Whipping Cords Will Serve You More Than Ropes Will Ever Do (Expanded Edition)", 
                        "The Idler Wheel...", album_name)) +  # Set plot title
    scale_x_continuous(limits = c(0, 1)) +  # Set x-axis limits
    scale_y_continuous(limits = c(0, 1))  # Set y-axis limits
    
  # Convert ggplot object to plotly
  p <- ggplotly(p, tooltip = "text")
  p <- p %>% layout(showlegend = FALSE)
  
  # Return the plot
  return(p)
}


# Define UI
ui <- fluidPage(
  titlePanel("Fiona Apple's Discography Analysis"),
  fluidRow(
    column(12, plotlyOutput("albumPlot", height = 500, width = 500))
  ),
  fluidRow(
    column(6, align = "right", actionButton("prevButton", "Previous")),
    column(6, align = "left", actionButton("nextButton", "Next"))
  )
)

# Define server logic
server <- function(input, output, session) {
  # Initialize index for album order
  album_index <- reactiveVal(1)
  
  # Update album index when previous button is clicked
  observeEvent(input$prevButton, {
    album_index(album_index() - 1)
    if (album_index() < 1) {
      album_index(length(album_order))
    }
  })
  
  # Update album index when next button is clicked
  observeEvent(input$nextButton, {
    album_index(album_index() + 1)
    if (album_index() > length(album_order)) {
      album_index(1)
    }
  })
  
  # Render plot for current album
  output$albumPlot <- renderPlotly({
    # Get current album name
    current_album <- album_order[album_index()]
    
    # Generate and display plot for current album
    plot_album(current_album)
  })
}

# Run the application
shinyApp(ui = ui, server = server)
```

-----------------------------------------------------------------------------

For this computational musicological analysis, I have chosen to focus on the discography of Fiona Apple (born 1977). Fiona burst onto the alternative jazz and pop scene in the late 90s as a teenager, captivating audiences with her aggressive piano ballads, emotionally charged vocal deliveries, and exceptional songwriting skills. Although I will not delve extensively into her songwriting prowess, it is worth acknowledging its significance in her musical identity.

I will be examining the evolution of Fiona Apple's sound over the years, spanning five full studio albums released between 1996 and 2020: "Tidal" (1996), "When The Pawn..." (1999), "Extraordinary Machine" (2005), "The Idler Wheel..." (2012), and "Fetch The Bolt Cutters" (2020). Despite the significant shifts in the musical landscape over this 24-year period, Fiona has consistently found her place within it, adapting her sound while retaining the essence of what makes a Fiona Apple song distinct.

For this analysis, I will focus solely on the songs from her albums, excluding any singles. This decision is reflective of Fiona Apple's discography, as she has released few standalone singles apart from her full album projects. I am interested in measuring the progression of each full project, as it encapsulates the evolution of Fiona's artistic vision and musical expression.

By examining these tracks and others across her discography, I aim to provide a comprehensive understanding of Fiona Apple's musical evolution and the broader context of her contributions to the alternative music scene.

### Radical Sensitivity: Visualized

```{r}
library(shiny)
library(shinyjs)
library(spotifyr)
library(tidyverse)
library(ggplot2)
library(plotly)

# Fetch the data from the Spotify API
corpus <- get_playlist_audio_features("", "7LDyKnV7gitLREBYl6kVLz")

# Specify the order of albums
album_order <- c("Tidal", "When The Pawn...", "Extraordinary Machine", 
                 "The Idler Wheel Is Wiser Than the Driver of the Screw and Whipping Cords Will Serve You More Than Ropes Will Ever Do (Expanded Edition)", 
                 "Fetch The Bolt Cutters")

# Define color palette
album_colors <- c("Tidal" = "blue", 
                  "When The Pawn..." = "red", 
                  "Extraordinary Machine" = "green", 
                  "The Idler Wheel Is Wiser Than the Driver of the Screw and Whipping Cords Will Serve You More Than Ropes Will Ever Do (Expanded Edition)" = "brown", 
                  "Fetch The Bolt Cutters" = "purple")

# Function to create a plot for each album
plot_album <- function(album_name) {
  # Filter data for the specified album
  album_data <- corpus %>% filter(track.album.name == album_name)
  
  # Create the plot
  p <- ggplot(album_data, aes(x = acousticness, y = valence, size = energy, 
                              text = paste("Song:", track.name, "<br>Album:", track.album.name, "<br>Energy:", energy))) +  # Define tooltip text
    geom_point(aes(color = album_name)) +  # Set color to album
    scale_color_manual(values = album_colors, guide = FALSE) +  # Remove legend for album name
    scale_size_continuous(range = c(3, 10), guide = FALSE) +  # Set point size without legend
    theme_minimal() +
    labs(title = ifelse(album_name == "The Idler Wheel Is Wiser Than the Driver of the Screw and Whipping Cords Will Serve You More Than Ropes Will Ever Do (Expanded Edition)", 
                        "The Idler Wheel...", album_name)) +  # Set plot title
    scale_x_continuous(limits = c(0, 1)) +  # Set x-axis limits
    scale_y_continuous(limits = c(0, 1))  # Set y-axis limits
    
  # Convert ggplot object to plotly
  p <- ggplotly(p, tooltip = "text")
  p <- p %>% layout(showlegend = FALSE)
  
  # Return the plot
  return(p)
}


# Define UI
ui <- fluidPage(
  titlePanel("Fiona Apple's Discography Analysis"),
  fluidRow(
    column(12, plotlyOutput("albumPlot", height = 500, width = 500))
  ),
  fluidRow(
    column(6, align = "right", actionButton("prevButton", "Previous")),
    column(6, align = "left", actionButton("nextButton", "Next"))
  )
)

# Define server logic
server <- function(input, output, session) {
  # Initialize index for album order
  album_index <- reactiveVal(1)
  
  # Update album index when previous button is clicked
  observeEvent(input$prevButton, {
    album_index(album_index() - 1)
    if (album_index() < 1) {
      album_index(length(album_order))
    }
  })
  
  # Update album index when next button is clicked
  observeEvent(input$nextButton, {
    album_index(album_index() + 1)
    if (album_index() > length(album_order)) {
      album_index(1)
    }
  })
  
  # Render plot for current album
  output$albumPlot <- renderPlotly({
    # Get current album name
    current_album <- album_order[album_index()]
    
    # Generate and display plot for current album
    plot_album(current_album)
  })
}

# Run the application
shinyApp(ui = ui, server = server)
```

------------------------------------------------------------------------

Fiona Apple's artistic evolution, analyzed through the lens of valence versus acousticness, reveals a nuanced progression across her discography.

In her debut album "Tidal," tracks like "Sleep to Dream" exhibit high valence and energy, with relatively low acousticness despite their acoustic roots. This juxtaposition suggests a youthful defiance and emotional intensity characteristic of her early work.

Transitioning to "When The Pawn...," we observe a departure from the negative correlation between acousticness and valence seen in her earlier album. "Paper Bag" stands out as a track with both high valence and acousticness, challenging the previous pattern and showcasing a more nuanced approach to instrumentation and emotional expression.

With "Extraordinary Machine," Fiona explores stripped-down arrangements and introspective melodies. The clustering of points on the left side of the plot indicates a preference for quieter moments, as evident in tracks like "Parting Gift," where acoustic instrumentation complements Fiona's haunting vocals, evoking a sense of introspection and longing.

"The Idler Wheel..." marks a further departure from Fiona's earlier sound, characterized by a shift towards slower, more introspective tracks. Despite the decrease in aggression, tracks like "Every Single Night" and "Werewolf" maintain a balance between acoustic elements and emotional depth.

In "Fetch The Bolt Cutters," Fiona defies conventional correlations between acousticness and valence. Here, tracks exhibit a diverse range of sonic textures and emotional tones, from the percussive chaos of "Cosmonauts" to the haunting beauty of "Shameika." This album showcases Fiona's willingness to experiment and challenge musical conventions, resulting in a rich and eclectic sonic landscape.

Overall, Fiona Apple's trajectory as an artist, as observed through the interplay of valence and acousticness in her music, reflects a journey marked by experimentation, evolution, and a steadfast commitment to artistic integrity.

### Chordogram

```{r}
library(tidyverse)
library(compmus)

circshift <- function(v, n) {
  if (n == 0) v else c(tail(v, n), head(v, -n))
}

#      C     C#    D     Eb    E     F     F#    G     Ab    A     Bb    B
major_chord <-
  c(   1,    0,    0,    0,    1,    0,    0,    1,    0,    0,    0,    0)
minor_chord <-
  c(   1,    0,    0,    1,    0,    0,    0,    1,    0,    0,    0,    0)
seventh_chord <-
  c(   1,    0,    0,    0,    1,    0,    0,    1,    0,    0,    1,    0)

major_key <-
  c(6.35, 2.23, 3.48, 2.33, 4.38, 4.09, 2.52, 5.19, 2.39, 3.66, 2.29, 2.88)
minor_key <-
  c(6.33, 2.68, 3.52, 5.38, 2.60, 3.53, 2.54, 4.75, 3.98, 2.69, 3.34, 3.17)

chord_templates <-
  tribble(
    ~name, ~template,
    "Gb:7", circshift(seventh_chord, 6),
    "Gb:maj", circshift(major_chord, 6),
    "Bb:min", circshift(minor_chord, 10),
    "Db:maj", circshift(major_chord, 1),
    "F:min", circshift(minor_chord, 5),
    "Ab:7", circshift(seventh_chord, 8),
    "Ab:maj", circshift(major_chord, 8),
    "C:min", circshift(minor_chord, 0),
    "Eb:7", circshift(seventh_chord, 3),
    "Eb:maj", circshift(major_chord, 3),
    "G:min", circshift(minor_chord, 7),
    "Bb:7", circshift(seventh_chord, 10),
    "Bb:maj", circshift(major_chord, 10),
    "D:min", circshift(minor_chord, 2),
    "F:7", circshift(seventh_chord, 5),
    "F:maj", circshift(major_chord, 5),
    "A:min", circshift(minor_chord, 9),
    "C:7", circshift(seventh_chord, 0),
    "C:maj", circshift(major_chord, 0),
    "E:min", circshift(minor_chord, 4),
    "G:7", circshift(seventh_chord, 7),
    "G:maj", circshift(major_chord, 7),
    "B:min", circshift(minor_chord, 11),
    "D:7", circshift(seventh_chord, 2),
    "D:maj", circshift(major_chord, 2),
    "F#:min", circshift(minor_chord, 6),
    "A:7", circshift(seventh_chord, 9),
    "A:maj", circshift(major_chord, 9),
    "C#:min", circshift(minor_chord, 1),
    "E:7", circshift(seventh_chord, 4),
    "E:maj", circshift(major_chord, 4),
    "G#:min", circshift(minor_chord, 8),
    "B:7", circshift(seventh_chord, 11),
    "B:maj", circshift(major_chord, 11),
    "D#:min", circshift(minor_chord, 3)
  )

key_templates <-
  tribble(
    ~name, ~template,
    "Gb:maj", circshift(major_key, 6),
    "Bb:min", circshift(minor_key, 10),
    "Db:maj", circshift(major_key, 1),
    "F:min", circshift(minor_key, 5),
    "Ab:maj", circshift(major_key, 8),
    "C:min", circshift(minor_key, 0),
    "Eb:maj", circshift(major_key, 3),
    "G:min", circshift(minor_key, 7),
    "Bb:maj", circshift(major_key, 10),
    "D:min", circshift(minor_key, 2),
    "F:maj", circshift(major_key, 5),
    "A:min", circshift(minor_key, 9),
    "C:maj", circshift(major_key, 0),
    "E:min", circshift(minor_key, 4),
    "G:maj", circshift(major_key, 7),
    "B:min", circshift(minor_key, 11),
    "D:maj", circshift(major_key, 2),
    "F#:min", circshift(minor_key, 6),
    "A:maj", circshift(major_key, 9),
    "C#:min", circshift(minor_key, 1),
    "E:maj", circshift(major_key, 4),
    "G#:min", circshift(minor_key, 8),
    "B:maj", circshift(major_key, 11),
    "D#:min", circshift(minor_key, 3)
  )

carrion <-
  get_tidy_audio_analysis("3pOxPpj3JhsmRAlaMrrxSy") |>
  compmus_align(bars, segments) |>
  select(bars) |>
  unnest(bars) |>
  mutate(
    pitches =
      map(segments,
        compmus_summarise, pitches,
        method = "mean", norm = "euclidean"
      )
  )

carrion |> 
  compmus_match_pitch_template(
    key_templates,         # Change to chord_templates if descired
    method = "euclidean",  # Try different distance metrics
    norm = "euclidean"     # Try different norms
  ) |>
ggplot(
  aes(x = start + duration / 2, width = duration, y = name, fill = d)
) +
  geom_tile() +
  scale_fill_viridis_c(guide = "none") +
  theme_minimal() +
  labs(x = "Time (s)", y = "") +
  geom_vline(xintercept = c(82, 114, 165), linetype = "dotted", color = "white", linewidth = 1.5)  
```

------------------------------------------------------------------------

For the visualisation of different chords I picked the final track of Fiona Apple’s debut album Tidal: Carrion. In this track Fiona alternates between a C minor and C major section. Indicated by the dotted lines you can see a clear switch on the chordogram. In "Carrion" by Fiona Apple, switching from a major to a minor chord signals a change in the story's mood. It's like the moment when the main character realizes the person they're attracted to isn't good for them. The music reflects this shift from hopeful to sad, matching how the character feels as they understand the relationship's problems. So, the minor chord captures the sadness and disappointment the character experiences, showing how the story moves from hope to a more somber realization.

### Self Similarity and Spectogram

```{r}
library(tidyverse)
library(spotifyr)
library(compmus)

fast_as_you_can <-
  get_tidy_audio_analysis("1UmmSKPdCtBuJsmbZG9G3u") |> # Change URI.
  compmus_align(bars, segments) |>                     # Change `bars`
  select(bars) |>                                      #   in all three
  unnest(bars) |>                                      #   of these lines.
  mutate(
    pitches =
      map(segments,
        compmus_summarise, pitches,
        method = "rms", norm = "euclidean"              # Change summary & norm.
      )
  ) |>
  mutate(
    timbre =
      map(segments,
        compmus_summarise, timbre,
        method = "rms", norm = "euclidean"              # Change summary & norm.
      )
  )

fast_as_you_can |>
  compmus_gather_timbre() |>
  ggplot(
    aes(
      x = start + duration / 2,
      width = duration,
      y = basis,
      fill = value
    )
  ) +
  geom_tile() +
  labs(x = "Time (s)", y = NULL, fill = "Magnitude", title = "Spectogram") +
  scale_fill_viridis_c() +                              
  theme_classic()

fast_as_you_can |>
  compmus_self_similarity(timbre, "cosine") |> 
  ggplot(
    aes(
      x = xstart + xduration / 2,
      width = xduration,
      y = ystart + yduration / 2,
      height = yduration,
      fill = d
    )
  ) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "Timbre Self Similarity Matrix")

fast_as_you_can |>
  compmus_self_similarity(pitches, "cosine") |> 
  ggplot(
    aes(
      x = xstart + xduration / 2,
      width = xduration,
      y = ystart + yduration / 2,
      height = yduration,
      fill = d
    )
  ) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(x = "", y = "", title = "Pitch Self Similarity Matrix")

```

------------------------------------------------------------------------

I have chosen my favorite track of Fiona Apple: Fast as you can for these visualizations. The reason why is because the structure of the track is very interesting. And while you may not see that very clearly in Plot 1: Spectrogram of Pitches and Plot 2: Self Similarity Matrix of Timbre. It is very evident in the third visualization. The song can be divided into 6 sections. The song goes from a verse to a chorus, to a breakdown (visible in yellow) to another verse and chorus which eventually concludes in a long outro.

### Classification & Clustering

```{r}
library(spotifyr)
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(gganimate)
library(compmus)
library(ggdendro)
library(heatmaply)
library(kknn)
library(ranger)

get_conf_mat <- function(fit) {
  outcome <- .get_tune_outcome_names(fit)
  fit |> 
    collect_predictions() |> 
    conf_mat(truth = outcome, estimate = .pred_class)
}  

get_pr <- function(fit) {
  fit |> 
    conf_mat_resampled() |> 
    group_by(Prediction) |> mutate(precision = Freq / sum(Freq)) |> 
    group_by(Truth) |> mutate(recall = Freq / sum(Freq)) |> 
    ungroup() |> filter(Prediction == Truth) |> 
    select(class = Prediction, precision, recall)
}  

top <- get_playlist_audio_features("top 10 fiona apple", "1uY52u1AGRYv2KkZEvHDou")
bottom <- get_playlist_audio_features("bottom 10 fiona apple", "24QEHRacIo9kz3PNkhYiHR")
topandbottom <-
  bind_rows(
    top |> mutate(playlist = "Top 10 Fiona Apple"),
    bottom |> mutate(playlist = "Bottom 10 Fiona Apple"),
  ) |> 
  add_audio_analysis()

topandbottom_features <-
  topandbottom |>  # For your portfolio, change this to the name of your corpus.
  mutate(
    playlist = factor(playlist),
    segments = map2(segments, key, compmus_c_transpose),
    pitches =
      map(
        segments,
        compmus_summarise, pitches,
        method = "mean", norm = "manhattan"
      ),
    timbre =
      map(
        segments,
        compmus_summarise, timbre,
        method = "mean",
      )
  ) |>
  mutate(pitches = map(pitches, compmus_normalise, "clr")) |>
  mutate_at(vars(pitches, timbre), map, bind_rows) |>
  unnest(cols = c(pitches, timbre))

view(topandbottom_features)

topandbottom_recipe <-
  recipe(
    playlist ~
      danceability +
      energy +
      loudness +
      speechiness +
      acousticness +
      instrumentalness +
      liveness +
      valence +
      tempo +
      duration +
      C + `C#|Db` + D + `D#|Eb` +
      E + `F` + `F#|Gb` + G +
      `G#|Ab` + A + `A#|Bb` + B +
      c01 + c02 + c03 + c04 + c05 + c06 +
      c07 + c08 + c09 + c10 + c11 + c12,
    data = topandbottom_features           # Use the same name as the previous block.
  ) |>
  step_center(all_predictors()) |>
  step_scale(all_predictors())      # Converts to z-scores.
  # step_range(all_predictors())    # Sets range to [0, 1].

topandbottom_cv <- topandbottom_features |> vfold_cv(5)

knn_model <-
  nearest_neighbor(neighbors = 1) |>
  set_mode("classification") |> 
  set_engine("kknn")

topandbottom_knn <- 
  workflow() |> 
  add_recipe(topandbottom_recipe) |> 
  add_model(knn_model) |> 
  fit_resamples(topandbottom_cv, control = control_resamples(save_pred = TRUE))

topandbottom_knn |> get_pr()

forest_model <-
  rand_forest() |>
  set_mode("classification") |> 
  set_engine("ranger", importance = "impurity")

topandbottom_forest <- 
  workflow() |> 
  add_recipe(topandbottom_recipe) |> 
  add_model(forest_model) |> 
  fit_resamples(
    topandbottom_cv, 
    control = control_resamples(save_pred = TRUE)
  )

topandbottom_forest |> get_pr()

workflow() |> 
  add_recipe(topandbottom_recipe) |> 
  add_model(forest_model) |> 
  fit(topandbottom_features) |> 
  pluck("fit", "fit", "fit") |>
  ranger::importance() |> 
  enframe() |> 
  mutate(name = fct_reorder(name, value)) |> 
  ggplot(aes(name, value)) + 
  geom_col() + 
  coord_flip() +
  theme_minimal() +
  labs(x = NULL, y = "Importance")

topandbottom_knn |> get_conf_mat() |> autoplot(type = "heatmap")

topandbottom_features |>
  ggplot(aes(x = acousticness, y = valence, colour = playlist, size = energy)) +
  geom_point(alpha = 0.8) +
  scale_color_viridis_d() +
  labs(
    x = "Acousticness",
    y = "Valence",
    size = "Energy",
    colour = "Playlist"
  )
```

------------------------------------------------------------------------

Efforts to use classification algorithms to tell apart successful and unsuccessful Fiona Apple songs didn't quite pan out. The plots and confusion matrices didn't show any clear differences, suggesting the features used might not be enough to spot what makes a song successful. It seems like we need better or more features to really figure out what sets apart the hits from the misses. So, we might need to tweak our approach or look for more factors to get a clearer picture of what makes Fiona Apple's songs succeed.

### Tempo Analysis

```{r}
library(shiny)
library(shinyjs)
library(spotifyr)
library(tidyverse)
library(ggplot2)
library(plotly)
library(compmus)

fayc <- get_tidy_audio_analysis("1UmmSKPdCtBuJsmbZG9G3u")

fayc |>
  tempogram(window_size = 8, hop_size = 1, cyclic = TRUE) |>
  ggplot(aes(x = time, y = bpm, fill = power)) +
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Tempo (BPM)", title = "Tempogram: Fast As You Can") +
  theme_classic()
```

------------------------------------------------------------------------

In analyzing the BPM of "Fast as You Can" by Fiona Apple, a noticeable dip in tempo occurs around 85 seconds, coinciding with a slower breakdown in the song. However, Apple quickly picks up the pace again, maintaining a dynamic rhythm throughout. This observation highlights the song's structural variation, with shifts in tempo adding to its overall musical complexity. Such insights into tempo dynamics offer a glimpse into how artists like Fiona Apple use pacing to craft an engaging song.

### Conclusions

```{r}
```

------------------------------------------------------------------------

Fiona Apple's musical journey from "Tidal" to "Fetch The Bolt Cutters" showcases her fearless experimentation and artistic growth. She defies genre boundaries, blending emotional depth with innovative instrumentation. While her chord changes and tempo shifts add layers to her storytelling, attempts to classify her songs reveal the complexity of her artistry. Fiona's uniqueness lies in her ability to balance vulnerability with innovation, creating music that resonates with authenticity. Whether it's her aggressive piano ballads or haunting vocal delivery, Fiona captivates audiences with her raw emotion and uncompromising vision. In a musical landscape driven by trends, Fiona Apple stands out as a true original, shaping her own path with every note she sings.
